---
title: "drones_cleaning"
author: "Paul Oldham"
date: "25 September 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document contains the code used to parse the Lens patent database results for a search for "drone" or "drones" in the full text and the title, abstract or claims. The tables generated are used in the inographic session of the WIPO patent analytics training.

Note that the data is limited using filter() to the period 1990 -2

```{r}
library(readr)
library(plyr)
library(dplyr)
library(tidyr)
library(stringr)
```


```{r}
drones_cleaned <- readr::read_csv("/Users/pauloldham17inch/Desktop/WIPO\ Training/dronesr/drones_fulltext/drones_fulltext_cleaned_5884.csv") %>% 
  oldhammisc::addcol(., "weight", 1) # note use of oldhammisc on github here.
```

```{r}
drones_cleaned <- dplyr::filter(drones_cleaned, publication_year >= 1990, publication_year <= 2015)
```

Generate applicants table

```{r}
organisations_cleaned <- dplyr::select(drones_cleaned, applicants_organisations_only, weight) %>% 
  tidyr::separate_rows(applicants_organisations_only, sep = ";") 

organisations_cleaned$applicants_organisations_only <-  stringr::str_trim(organisations_cleaned$applicants_organisations_only, side = "both")

organisations_count <- dplyr::count(organisations_cleaned, applicants_organisations_only, wt = weight) %>% 
  na.omit() %>% 
  dplyr::arrange(desc(n))
readr::write_csv(organisations_count, "drones_organisations_count_fulltext.csv")
```

Inventors

```{r}
inventors_cleaned <- select(drones_cleaned, inventors_useme, weight) %>% 
  tidyr::separate_rows(inventors_useme, sep = ";") 
inventors_cleaned$inventors_useme <-  stringr::str_trim(inventors_cleaned$inventors_useme, side = "both")
inventors_count <- count(inventors_cleaned, inventors_useme, wt = weight) %>% 
  na.omit() %>% 
  arrange(desc(n))
readr::write_csv(inventors_count, "drones_inventors_count_fulltext.csv")
```

IPCs

```{r}
ipc_short <- select(drones_cleaned, ipc_short, weight) %>% 
  tidyr::separate_rows(ipc_short, sep = ";") 
ipc_short$ipc_short <-  stringr::str_trim(ipc_short$ipc_short, side = "both")
ipc_short_count <- count(ipc_short, ipc_short, wt = weight) %>% 
  na.omit() %>% 
  arrange(desc(n))
readr::write_csv(ipc_short_count, "drones_ipc_short_count_fulltext.csv")
```

phrases

```{r}
phrases <- select(drones_cleaned, title_phrases, weight) %>% 
  tidyr::separate_rows(title_phrases, sep = ";") 
phrases$title_phrases <-  stringr::str_trim(phrases$title_phrases, side = "both")
phrases_count <- count(phrases, title_phrases, wt = weight) %>% 
  na.omit() %>% 
  arrange(desc(n))
readr::write_csv(phrases, "drones_phrases_fulltext.csv")
```

Needs to be the top 10?

Trends - Publication count by country

Needs to be in format for infor.gram

```{r}
country_totals <- select(drones_cleaned, publication_country_name, publication_year, weight) %>%
  count(publication_country_name, publication_year, wt = weight) %>% # note n
  spread(publication_year, n) # note nn
country_totals
readr::write_csv(country_totals, "drones_country_totals_fulltext.csv")
```

Filtering Main data frame

```{r}
tac <- dplyr::filter(drones_cleaned, title_abstract_claims == "tac")
fulltext <- dplyr::filter(drones_cleaned, title_abstract_claims == "full_text")
```

###Title Abstract Claims Dataset

```{r}
drones_tac_cleaned <- readr::read_csv("/Users/pauloldham17inch/Desktop/WIPO\ Training/dronesr/drones_tac/drones_tac_1606_cleaned.csv") %>% 
  oldhammisc::addcol(., "weight", 1)
drones_tac_cleaned <- dplyr::filter(drones_tac_cleaned, publication_year >= 1990, publication_year <= 2015)
```

###Generate applicants TAC table

```{r}
organisations_tac <- select(drones_tac_cleaned, applicants_organisations_only, weight) %>% 
  tidyr::separate_rows(applicants_organisations_only, sep = ";") 
organisations_tac$applicants_organisations_only <-  stringr::str_trim(organisations_tac$applicants_organisations_only, side = "both")
organisations_tac_count <- count(organisations_tac, applicants_organisations_only, wt = weight) %>% 
  na.omit() %>% 
  arrange(desc(n))
readr::write_csv(organisations_tac_count, "drones_organisations_count_tac.csv")
```

###Inventors TAC

```{r}
inventors_tac <- select(drones_tac_cleaned, inventors_useme, weight) %>% 
  tidyr::separate_rows(inventors_useme, sep = ";") 
inventors_tac$inventors_useme <-  stringr::str_trim(inventors_tac$inventors_useme, side = "both")
inventors_tac_count <- count(inventors_tac, inventors_useme, wt = weight) %>% 
  na.omit() %>% 
  arrange(desc(n))
readr::write_csv(inventors_tac_count, "drones_inventors_count_tac.csv")
```

###IPC TAC
```{r}
ipc_tac_short <- select(drones_tac_cleaned, ipc_short, weight) %>% 
  tidyr::separate_rows(ipc_short, sep = ";") 
ipc_tac_short$ipc_short <-  stringr::str_trim(ipc_tac_short$ipc_short, side = "both")
ipc_tac_short_count <- count(ipc_tac_short, ipc_short, wt = weight) %>% 
  na.omit() %>% 
  arrange(desc(n))
readr::write_csv(ipc_tac_short_count, "drones_ipc_short_count_tac.csv")
```

###phrases TAC

```{r}
phrases_tac <- select(drones_tac_cleaned, title_phrases, weight) %>% 
  tidyr::separate_rows(title_phrases, sep = ";") 
phrases_tac$title_phrases <-  stringr::str_trim(phrases_tac$title_phrases, side = "both")
phrases_tac_count <- count(phrases_tac, title_phrases, wt = weight) %>% 
  na.omit() %>% 
  arrange(desc(n))
readr::write_csv(phrases_tac_count, "drones_phrases_count_tac.csv")
```

```{r}
country_tac_totals <- select(drones_tac_cleaned, publication_country_name, publication_year, weight) %>%
  count(publication_country_name, publication_year, wt = weight) %>% # note n
  spread(publication_year, n) # note nn
country_tac_totals
readr::write_csv(country_tac_totals, "drones_country_totals_tac.csv")
```


```{r}
#testfunct <- function(x, col = "", count = ""){
#  df <- dplyr::select_(x, col, count) %>% 
#    tidyr::separate_rows_(col, sep = ";")
#  df[col] <- stringr::str_trim(df[col], side = "both")
#  df #behaving rather weirdly by producing concatenated tibble
#}
#testme <- testfunct(drones_cleaned, "inventors_useme", "weight") 
```
